---
title: "Heatmaps"
author: "Joyce Robbins"
date: "8/17/2017"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, 
                      warning = FALSE, message = FALSE)
```


```{r, echo = FALSE}
library(ggplot2)
library(gridExtra)
```

### geom_tile with numerical data, compare to geom_point  
```{r}
x <- 1:3
y <- c(5, 2, 7)
df <- data.frame(x, y)
g1 <- ggplot(df, aes(x, y)) + geom_point()
g2 <- ggplot(df, aes(x, y)) + geom_tile()
grid.arrange(g1, g2, nrow = 1)
```

### geom_tile with variable w, h
```{r}
df$w <- c(1.4, .4, .2)
df$h <- c(.5, 1.3, .8)
ggplot(df, aes(x, y)) + geom_tile(aes(width = w, height = h))
```

### geom_rect uses (xmin, xmax, ymin, ymax) instead
```{r}
xmin <- 1:3
xmax <- 2:4
ymin <- c(5, 2, 7)
ymax <- c(6, 3, 8)
df <- data.frame(xmin, xmax, ymin, ymax)
ggplot(df, aes(xmin = xmin, xmax = xmax, ymin = ymin,
               ymax = ymax)) + geom_rect()


```

### force squares
```{r}
ggplot(df, aes(xmin = xmin, xmax = xmax, ymin = ymin,
               ymax = ymax)) + geom_rect() + coord_fixed()
```

### geom_raster -- same as geom_tile w/ uniform w, h & FASTER
```{r}
x <- 1:3
y <- c(5, 2, 7)
df <- data.frame(x, y)

ggplot(df, aes(x,y)) + geom_raster()
```

### change x & y: what happens to the size of the tiles?
```{r}
x <- 1:2
y <- c(5, 2)
df <- data.frame(x, y)
ggplot(df, aes(x, y)) + geom_raster()
```

### change x & y again
```{r}
x <- 1:4
y <- c(5, 2, 10, 7)
df <- data.frame(x, y)
ggplot(df, aes(x, y)) + geom_raster()
```

## Practice using geom_tile(), geom_rect(), geom_raster()

## Building the heatmap

### complete set of (x, y) pairs
```{r}
x <- c(1, 1, 1, 2, 2, 2, 3, 3, 3)
y <- c(1, 2, 3, 1, 2, 3, 1, 2, 3)
df <- data.frame(x, y)
ggplot(df, aes(x, y)) + geom_raster()
```

### add color
```{r}
set.seed(2017)
df$z <- sample(9)
ggplot(df, aes(x, y)) + geom_raster(aes(fill = z))
```

### What if z is categorical?
```{r}
df$z <- c("A", "B", "C", "B", "A", "A", "B", "C", "B")
ggplot(df, aes(x, y)) + geom_raster(aes(fill = z))
```

### What is z is discrete (numerical)?
```{r}
df$z <- sample(3, 9, replace = TRUE)
ggplot(df, aes(x, y)) + geom_raster(aes(fill = z))
df$z <- factor(df$z)
ggplot(df, aes(x, y)) + geom_raster(aes(fill = z))

```

### clean it up
```{r}
theme_heat <- theme_classic(14) +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank())

ggplot(df, aes(x, y)) + geom_raster(aes(fill = z)) + 
  theme_heat
```

### add coord_fixed
```{r}
ggplot(df, aes(x, y)) + geom_raster(aes(fill = z)) + 
  coord_fixed() + theme_heat
```

### add white border (doesn't work with geom_raster)  

```{r}
ggplot(df, aes(x, y)) + 
  geom_tile(aes(fill = z), color = "white") + 
  coord_fixed() + theme_heat
```
